Declarative Deployment Reconfiguration of Pulverised Collective Adaptive Systems at Runtime \thanks{Identify applicable funding agency here. If none, delete this.}

### 2023-05-16 paper-2023-acsos-pulverization.tex
Abstract: + examples of domains

\begin{abstract}
In recent years,
the infrastructure supporting the execution of situated distributed computations
evolved with a fast pace.
%
Modern collective adaptive applications are designed to be executed on very diverse devices
and to be deployed on infrastructures composed of devices ranging from cloud servers to wearable devices,
constituting together a cloud-edge continuum.
%
The availability of such an infrastructure opens to better resource utilisation and performance,
but, contextually, introduces new challenges to software designers,
as applications must be conceived to be able to adapt their deployment to changing conditions.
%
In this paper,
we introduce a practical framework for the development of collective adaptive systems
based on the concept of \emph{pulverisation},
meant to neatly separate business logic and deployment concerns,
allowing applications to be defined independent of the infrastructure they will execute upon.
%
The framework is based on a domain-specific language capturing,
in a declarative fashion,
pulverised application components, device capabilities, resource allocation, and (runtime re-) configuration policies.
%
The framework, implemented in Kotlin multiplatform and available as open source,
is then evaluated in a small-scale real-world demo and in a city-scale simulated scenario.
\end{abstract}

@@

\section{A framework for declarative runtime reconfiguration}\label{sec:contribution}

@@

\subsection{Domain-specific languages for pulverisation}

In principle, there are several possibilities to implement a pulverisation configurator and runtime:
as a library
configured and executed by a target programming language,
as a framework/runtime configured by means of configuration files and with its own runtime,
or as a \ac{dsl} configured with its own custom syntax.
%
These approaches have different trade-offs.
%
A (well designed) library can be flexible and easier to adopt
as it does not require to learn new languages;
but it will be usable only in the language/runtime it has been designed for,
and will require its own framework/runtime to be executed
(or appropriate bindings for existing ones).
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but it usually dictates network protocols and features limited flexibility,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
Finally, a \ac{dsl} can be designed to be as expressive as needed,
but it has high maintenance cost (as the language maintenance costs stack upon those of the library/API) and, potentially,
a steep learning curve due to the need of learning a new (custom) language.
%
Recent evolution in programming languages,

such as Kotlin, Ruby, and Scala~\cite{Riti2018}
however,
a trend has emerged that
TODO: recently, internal DSLs
adopted by gradle

From the previous section,
we can identify two main phases in the pulverisation process:
\begin{enumerate*}[label=\it{(\roman*)}]
  \item the definition of the available devices and their capabilities (infrastructure); and
  \item the allocation of the pulverised components to them (runtime).
\end{enumerate*}
%
We decided to capture these phases into two distinct \acp{dsl}.

>>>>

\begin{abstract}
In recent years,
the infrastructure supporting the execution of situated distributed computations
evolved with a fast pace.
%
Modern collective adaptive applications -- as found in the Internet of Things, swarm robotics, and social computing -- are designed to be executed on very diverse devices
and to be deployed on infrastructures composed of devices ranging from cloud servers to wearable devices,
constituting together a cloud-edge continuum.
%
The availability of such an infrastructure opens to better resource utilisation and performance,
but, contextually, introduces new challenges to software designers,
as applications must be conceived to be able to adapt their deployment to changing conditions.
%
In this paper,
we introduce a practical framework for the development of collective adaptive systems
based on the concept of \emph{pulverisation},
meant to neatly separate business logic and deployment concerns,
allowing applications to be defined independent of the infrastructure they will execute upon.
%
The framework is based on a domain-specific language capturing,
in a declarative fashion,
pulverised application components, device capabilities, resource allocation, and (runtime re-) configuration policies.
%
The framework, implemented in Kotlin multiplatform and available as open source,
is then evaluated in a small-scale real-world demo and in a city-scale simulated scenario.
\end{abstract}

@@

\section{Background}\label{sec:background}

\subsection{Pulverisation}

@@

\subsection{Domain-specific languages}

In principle, there are several possibilities to implement a pulverisation configurator and runtime:
as a library
configured by a program written in a general-purpose language,
as a framework/runtime configured by means of configuration files,
or as a \ac{dsl} configured with its own custom syntax.
%
These approaches have different trade-offs.
%
A (well designed) library can be flexible as the host language can be used to express peculiar configurations
and easier to adopt to users acquainted with the host language;
but its configuration may quickly become imperative and harder to maintain as complexity grows.
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but its flexibility is limited,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
Finally, a \ac{dsl} can be designed to be as expressive as needed,
but it has high maintenance cost (as the language maintenance costs stack upon those of the library/API) and, potentially,
a steep learning curve due to the need of learning a new (custom) language.
%
Recent evolution in programming languages,
however,
open to an additional strategy:
the construction of \emph{internal} \acp{dsl},
hybrids between libraries and stand-alone (external) \acp{dsl}.
%
In fact, modern languages such as Kotlin, Groovy, Ruby, and Scala~\cite{Riti2018}
provide specific syntactic features to enable the construction of APIs whose ergonomy is akin to the one of a dedicated language,
but that are valid fragments in the host language.
%
Although internal \acp{dsl} are de facto libraries in the host programming language
(there is no clear boundary defining when a library becomes a \ac{dsl}\footnote{\url{https://archive.is/wip/xAeiX}}),
from a practical perspective
they allow for great flexibility and ergonomy
(although not total as the one of a custom \ac{dsl}, as they are subject to the syntactic constraints of the host language)
while retaining a reduced maintenance cost
(as the host language ecosystem and tooling can be reused directly)
and a gentle learning curve
(as the syntax will be largely familiar to the host language users).
%
The internal DSL approach has been successfully applied in several domains,
spanning from
build systems\footnote{\url{https://archive.is/5xtaN}}
and ontologies~\cite{DBLP:journals/jossw/Balhoff16} to
hardware design~\cite{DBLP:journals/trets/SerreP20} to
logic~\cite{2pkt-swx16} and aggregate programming~\cite{DBLP:journals/softx/CasadeiVAP22}.

\section{A \ac{dsl} for runtime reconfiguration of pulverised systems}\label{sec:contribution}

From the previous section,
we can identify two main phases in the pulverisation process:
\begin{enumerate*}[label=\it{(\roman*)}]
  \item the definition of the available devices and their capabilities (\emph{infrastructure}); and
  \item the allocation of the pulverised components to them (\emph{runtime}).
\end{enumerate*}
%
Coherently with the single responsibility principle,
we decided to capture these phases into two distinct \acp{dsl}.
